From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-07 Pi+/paper/discussion.tex
Add some ideas for discussion

Pi types -- Natural number -- Finite sets
1-combinator -- Generators of Sn -- 1-paths
2-combinator -- Relations of Sn -- 2-paths

In HoTT, univalence characterises the path type in the universe as equivalences of types. The map $\term{idtoeqv} : A
\id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term $\term{ua} : A \eqv B \to A \id_{\UU}
B$, its computation rule $\term{ua-\beta} : (e : A \eqv B) \to \term{idtoeqv}(\term{ua}(e)) \id e$, and its
extensionality rule $\term{ua-\eta} : (p : A \id_{\UU} B) \to p == \term{ua}(\term{idtoeqv(p)})$ are generally added as
postulates when formalising in Agda. Together, $\term{ua}$ and $\term{ua}-\beta$ give the full univalence axiom $(A \eqv
B) \eqv (A \id_{\UU} B)$.

@@

Free commutative monoids in type theory have been studied in~\cite{gylterudMultisetsTypeTheory2020}, and using HoTT
in~\cite{choudhuryFinitemultisetConstructionHoTT2019}. The free symmetric monoidal groupoid $\FSM[A]$ over a groupoid
$A$ can be given by $\dsum{X:\UFin}{A^{X}}$, or it can be presented as an algebraic 2-theory using 1-HITs. These HITs
and the proof of their universal property have been considered
in~\cite*{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}.

@@

>>>>

\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua-\beta} : (e : A \eqv B) \to
    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua-\eta} : (p : A \id_{\UU} B) \to p ==
    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
$\term{ua}-\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.

@@

Free commutative monoids in type theory have been studied in~\cite{gylterudMultisetsTypeTheory2020}, and using HoTT
in~\cite{choudhuryFinitemultisetConstructionHoTT2019}. The free symmetric monoidal groupoid $\FSM[A]$ over a groupoid
$A$ can be given by $\dsum*{X:\UFin}{A^{X}}$, or it can be presented as an algebraic 2-theory using 1-HITs. These HITs
and the proof of their universal property have been considered
in~\cite*{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}.

@@

\begin{center}
    \begin{tabular}{c c c}
        \hline
        $\PiLang$     & $\Sn$               & $\UFin$     \\
        \hline
        types         & Natural numbers     & Finite sets \\
        \hline
        1-combinators & Generators of $\Sn$ & 1-paths     \\
        \hline
        2-combinators & Relations of $\Sn$  & 2-paths     \\
        \hline
    \end{tabular}
\end{center}

\begin{center}
    \begin{tabular}{c c c}
        \hline
        Logic                              & Program                       & Algebra                     \\
        \hline
        Propositions                       & Types                         & Objects                     \\
        \hline
        Proofs                             & Terms                         & Morphisms                   \\
        \hline
        $\blank$                           & Equations                     & Homotopies                  \\
        \hline
        Intuitionistic propositional logic & Simply typed Lambda Calculus  & Cartesian Closed Categories \\
        \hline
        Superstructural Reversible logic   & Reversible Language $\PiLang$ & Symmetric Rig Groupoids     \\
        \hline
        Connectives                        & Type constructors             & Bifunctors                  \\
        \hline
        ?                                  & Type normalization            & ?                           \\
        \hline
        ?                                  & Program normalization         & ?                           \\
        \hline
        ?                                  & Soundness of 2-combinators    & ?                           \\
    \end{tabular}
\end{center}

\begin{center}
    \begin{tabular}{c c c}
        \hline
        Reversible Logic             & Reversible Programming                  & Algebra                  \\
        \hline
        Free commutative rig         & Free symmetric rig groupoid             & $\UFin$                  \\
        \hline
        Propositions                 & Types                                   & Objects                  \\
        \hline
        Entailment?                  & Reversible programs                     & Isomorphisms             \\
        \hline
        $\blank$                     & 2-combinators                           & Homotopies               \\
        \hline
        Conjunction?                 & $\times$                                & $\otimes$                \\
        \hline
        Disjunction?                 & $+$                                     & $\oplus$                 \\
        \hline
        $A \land B \vdash B \land A$ & $\swapt$                                & $\mathcal{B}_{A,B}$      \\
        \hline
        True proposition             & $\onet$                                 & $F_1$                    \\
        \hline
        Decidability                 & Program Synthesis                       & Reachability in a graph? \\
        \hline
        Soundness/Completeness       & Soundness/Completeness of 1-combinators & ?                        \\
    \end{tabular}
\end{center}
